Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations

### 2021-09-15 paper/latex/NetworkNeuroscience/rmt.tex
update

\input{preamble.tex}

\title{Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations}
\author{Derek M. Berger \and Gurpreet S. Matharoo \and Jacob Levman}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
    Previous studies have investigated the potential of using analytic techniques from Random Matrix
    Theory (RMT) to investigate magnetic resonance imaging (MRI) data. We assess the potential
    application of RMT-based features for the analysis of functional MRI (fMRI) across diverse
    datasets. As novel contributions, we (1) assess the potential for RMT-inspired, whole-brain
    features extracted from voxel-wise functional connectivity to contain information useful for
    classifying between various psychological processes, (2) assess these features’
    predictive—rather than explanatory—value, (3) investigate the effect of varying RMT analysis
    methods on the robustness of study findings, and (4) make code publicly available to extract
    these features from a wide variety of data. We find preliminary evidence suggesting that
    RMT-inspired features may have unique potential in analyses of fMRI functional connectivity.

    \subparagraph{Keywords:} feature extraction, functional connectivity, functional magnetic
    resonance imaging, random matrix theory, voxelwise connectivity, predictive analysis

    \subparagraph{Authors}
    Derek Berger is with the department of Computer Science, St. Francis Xavier University, Antigonish
    NS B2G 2W5, Canada (e-mail: dberger@stfx.ca).

    Gurpreet Matharoo is with ACENET, and the Department of Physics, St. Francis Xavier University,
    Antigonish NS B2G 2W5, Canada (e-mail: gmatharo@stfx.ca).

    Jacob Levman is with the Department of Computer Science, St. Francis Xavier University,
    Antigonish NS B2G 2W5, Canada (e-mail: jlevman@stfx.ca).
\end{abstract}

@@

Whether in the presence of experimental stimuli, or the relative absence, as in a resting state,
complex functional connectivity networks are ubiquitous
(\cite{bucknerBrainDefaultNetwork2008,foxCoverHumanBrain2005,gonzalez-castilloTaskbasedDynamicFunctional2018,hermundstadStructuralFoundationsRestingstate2013}).
This complexity suggests fMRI is a candidate to be studied using Random Matrix Theory (RMT), a set
of mathematical tools originally developed some 50 years ago to solve complex problems in nuclear
physics (\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}). In various
physical systems where nuclei can have different possible states, precisely modeling the entirety of
the interactions between nuclei is often intractable. For such large, complex systems, the
eigenvalues (or spectra) efficiently summarize the totality of the interactions between components
of the system, and RMT describes the expected behaviour of such eigenvalues. By analysing the
spectrum statistically, complex systems can be investigated by comparison to the universal
properties predicted by RMT.

\begin{sloppypar}
These universal properties have been observed in diverse phenomena. In small-scale systems, RMT
universalities have been observed in quantum chaotic systems, complex nuclei, atoms, molecules and
disordered mesoscopic systems \cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004,brodyRandommatrixPhysicsSpectrum1981,beenakkerRandommatrixTheoryQuantum1997,bohigasHigherOrderCorrelationsSpectra1985,wintgenLevelStatisticsQuantized1988,pandeySkewOrthogonalPolynomialsUniversality2001}, and at larger scales, RMT has been applied to atmospheric
physics (\cite{santhanamStatisticsAtmosphericCorrelations2001}, stock cross-correlations \cite{plerouRandomMatrixApproach2002} , social networks \cite{jalanUncoveringRandomnessSuccess2014}, random networks \cite{bandyopadhyayUniversalityComplexNetworks2007},
network-formation in liquids (\cite{sastrySpectralStatisticsInstantaneous2001,matharooSpectralStatisticsQuenched2009})  and amorphous clusters (\cite{sarkarUniversalityVibrationalSpectra2004,matharooVibrationalSpectraAmorphous2005,matharooUniversalityVibrationalSpectra2008}). Within biological systems,
RMT has also been used to successfully model aspects of  amino acid functional relationships (\cite{bhadolaTargetingFunctionalMotifs2016}),
synchrony in epileptic seizures (\cite{osorioPhasesynchronizationRandommatrixBased2011}), and in protein-protein interactions both in different species (\cite{agrawalQuantifyingRandomnessProtein2014}) and breast cancer (\cite{raiRandomnessPreservedPatterns2015}).
\end{sloppypar}

In recent years, RMT analyses have been applied to study brain functioning.  The earliest  study
demonstrated that spectra of the correlations between electroencephalographic (EEG) signals closely
resemble those of the Gaussian Orthogonal Ensemble (GOE) (\cite{sebaRandomMatrixAnalysis2003}).
Somewhat more recently, RMT was used to evaluate the quality of whole brain features extracted from
fMRI data (\cite{voultsidouFeatureEvaluationFMRI2007,verganiRestingStateFMRI2019}). RMT has also
been used in diffusion MRI to aid in the selection of the number of components to employ in
principal-component reduction analysis and denoising
(\cite{veraartDenoisingDiffusionMRI2016,verganiRestingStateFMRI2019,ulfarssonDimensionEstimationNoisy2008}).

Finally, RMT has been used in ROI-based fMRI functional-connectivity studies to investigate
differences between rest and task states (\cite{wangSpectralPropertiesTemporal2015}), between
subjects with and without ADHD (\cite{wangRandomMatrixTheory2016}), and between pain and non-pain
states (\cite{matharooSpontaneousBackpainAlters2020}). Across these three studies, the spectra of
resting or low-attention states exhibited properties closest to the GOE.

@@

However, since previous studies employing RMT for feature extraction
(\cite{wangRandomMatrixTheory2016,matharooSpontaneousBackpainAlters2020}) have partially interpreted
their findings with respect to attentional processes, we have elected to use datasets that involve
attention in some way. This includes measures of attention or vigilance, or comparisons between a
condition involving more attention and focus (e.g. a task state) or conditions which should involve
considerably less attention or focus (e.g. resting-state).

@@

Dataset scan parameters and acquisition details are summarized in Table I, and sample and subgroup
sizes are summarized in Table II. For all datasets, we use the raw (unpreprocessed) scans, and
perform our own preprocessing (detailed later). Given the complexity of these datasets, we highlight
below only those aspects relevant to the current paper.

@@

\subsubsection{Dataset 1}
Dataset 1 (referred to in Fig. and tables with the label ``PSYCH'') has detailed and extensive
information available in \cite{gorgolewskiHighResolution7Tesla2015}. Subjects completed a battery of
psychological measurements, and various resting-state fMRI (rs-fMRI) images were obtained. Only the
whole-brain 7-Tesla rs-fMRI scans were used for the current paper, to make the results comparable to
the other included datasets. For the purpose of keeping with the ``attention'' theme of the analyses
in this paper, three pairings of subgroups were constructed via z-score splits of the various
sum-scores of scale items involving concepts closely related to attention.

\paragraph{Trait Attention}
The PANAS-X (\cite{watsonPANASXManualPositive1994}) is a 9-point Likert-type self-report measure
that queries the extent to which subjects agree particular affective terms have applied to them over
a recent time period (e.g. a few weeks). An ad hoc (i.e. face-valid) sum-score was constructed from
the scale items: ``attentive'', ``alert'', ``interested'', ``concentrating'', and reverse-scored items
``sluggish'', ``tired'', ``sleepy'', ``drowsy''. Subjects with a z-score greater / less than zero on this
sum-score were placed in the high / low ``trait attention'' category

\paragraph{Task Attention}
The Conjunctive Continuous Performance Task (CCPT;
\cite{shalevConjunctiveContinuousPerformance2011}) is a behavioural measure of sustained attention
where subjects must react only if they see a red square among other different coloured figures. For
this paper, a sum score was constructed by subtracting the number of false positive and false
negative responses from the number of success responses, and subjects with a z-score greater / less
than zero on this scale were placed in high / low ``task attention'' category.

\paragraph{Vigilance}
The Mini New York cognition questionnaire assesses various elements of the subject's cognitive
state, e.g. ``my thoughts involved other people'', ``I thought about something positive'' (translated
from German; \cite{gorgolewskiCorrespondenceIndividualDifferences2014}). One question assesses
alertness (``I was fully awake'' / ``war ich vollkommen wach'') and was used to create high / low
``vigilance'' conditions in the same manner as the previous two splits.

\subsubsection{Dataset 2}
Dataset 2 (referred to in Fig. and tables with the label ``OSTEO'') is detailed in
\cite{tetreaultBrainConnectivityPredicts2016}. Subjects included healthy controls (``nopain''
condition) and individuals with knee osteoarthritis (``pain'' condition) treated for 2 weeks with
either placebo (``pain'' condition) or duloxetine (``duloxetine'' condition). An additional grouping was
also created with both the placebo and duloxetine data (``allpain'' condition), allowing for six
subgroup comparisons. All subjects were imaged with whole-brain rs-fMRI. Scan parameters are
documented in Table I. The use of rs-fMRI and RMT for examining pain makes this dataset
complementary to \cite{matharooSpontaneousBackpainAlters2020}.

\subsubsection{Dataset 3}
Dataset 3  (referred to here with the label ``REFLECT'', due to the ``reflective” nature of the
cognitive tasks) is described in full detail by \cite{dupreMultiechoFMRIReplication2016}. Both
multi-echo rs- and task-fMRI scans were conducted with scan parameters summarized in Table II. Tasks
involved viewing various affective images (see \cite{sprengPatternsBrainActivity2010} for details)
and performing tasks related to autobiographical memory, prospection, or theory-of-mind reasoning.
Continuing with the theme of broadly comparing high versus low attention throughout this paper, the
different tasks were simply treated generically as ``attending'', in contrast to the non-attentive
resting-state scans. This resulted in four task state scans and one resting-state scan per subject.

In order to combine the multi-echo data into a single, usable matrix, two methods were employed. In
the ``summed'' condition, echoes were combined via summation, a method which has value in providing an
image with overall increased contrast (\cite{kunduMultiechoFMRIReview2017}). In the ``interleave''
condition, echoes were interleaved at each voxel: that is, if we denote the voxel's ith signal value
from echo j as , then the interleaved voxel of the new image has the time-series \([t_{1,1},
t_{2,1}, t_{3,1}, \dots, t_{1,i}, t_{2,i}, t_{3,i}, \dots, t_{1,n}, t_{2,n}, t_{3,n}]\).

@@

\subsubsection{Dataset 4}
Dataset 4 (referred to in Fig. and tables with the label ``PARK'' or ``PARKINSONS'') is documented and
described in detail in \cite{madhyasthaDynamicConnectivityRest2015}. Scan parameters are documented
in Table I. Subjects with non-demented Parkinson’s Disease (PD) and healthy controls performed a
number of repetitions of the Attention Network Task (ANT)
(\cite{fanActivationAttentionalNetworks2005}). Subgroup details are available in Table I. As the ANT
task demands high attention, it is not as straightforward to create ``high'' or ``low'' attention
groups. Instead, since attention presumably is high and varies little in this dataset, and subgroups
instead differ by virtue of the presence or absence of PD, any results here serve as a useful
comparison point for the results from studies that should clearly involve attentional differences.
This dataset also included extensively preprocessed scans, which were included in analyses to
examine the effects of preprocessing on the extracted RMT features.

\subsubsection{Dataset 5}
Dataset 5 (referred to in Fig. and tables with the label ``LEARN'') is documented and described in
detail in \cite{schapiroHumanHippocampalReplay2018}. In this study, both rs-fMRI and task fMRI scans
were available for all subjects, allowing for within-subject comparisons. The complex task involved
multiple phases, wherein subjects were exposed to named, black-and-white outlined satellite images
related by family-resemblance of the components (see Fig. 1 of
\cite{schapiroHumanHippocampalReplay2018}). After this learning task, subjects were tested on their
memory of the various learned features. Scans corresponding to both the initial learning and
subsequent testing phase were included in the high-attention ``task'' subgroup for the purposes of the
analysis of this paper. Resting state scans were included in the low-attention ``rest'' subgroup.

\subsection{Preprocessing}
Some consideration should be given to whether it is appropriate to apply standard preprocessing
steps before extracting RMT features, as preprocessing steps should be developed, tested, and
employed always with respect to a set of inferential goals and a specific analytic framework.
However, these preprocessing steps are increasingly recognized to have complex effects on fMRI
analyses (\cite{parkerBenefitSliceTiming2019}). For example, spatial smoothing has clear benefits
when employing a general linear model
(\cite{miklEffectsSpatialSmoothing2008,alakorkkoEffectsSpatialSmoothing2017,worsleyAnalysisFMRITimeSeries1995}),
but is less motivated in functional network analyses with small ROIs, where the effect of smoothing
can be considerable
(\cite{alakorkkoEffectsSpatialSmoothing2017,goelmanMaximizingNegativeCorrelations2014,trianaEffectsSpatialSmoothing2020},
with these effects interacting complexly with factors like field strength and image resolution
(\cite{triantafyllouEffectSpatialSmoothing2006}). Similarly, image registration is usually required
to allow comparisons across (or even within) subjects, but the choice of registration strategy can
have dramatic effects on functional network analyses (\cite{magalhaesImpactNormalizationSegmentation2014}).

@@

For this paper’s analyses, brain extraction is necessary to limit analyses to the appropriate
voxels, motion correction is necessary to allow proper brain extraction, and both motion-correction
and slice-time correction simplify the interpretation of voxel-voxel correlations across voxel
locations. Whereas physiological noise regression and bias field correction would perhaps be
appropriate here, there is insufficient data to perform these steps for all datasets (Table II).
Thus, motion-correction was performed with FSL's MCFLIRT
(\cite{jenkinsonImprovedOptimizationRobust2002}), slicetime correction with the default method in
FSL, and brain extraction was performed with BET (\cite{smithFastRobustAutomated2002}). Due to
differences in the size of the REFLECT data, the higher-resolution resting-state data was
interpolated to have the same spatial dimensions as the task data.

@@

Two classes of features were examined in this study: those that require unfolding, and those that do
not. In order to compare the observed eigenvalue distribution to some theoretical predictions, the
eigenvalues must be unfolded (\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}).
The details and motivation behind the unfolding procedure are well documented in
\cite{guhrRandommatrixTheoriesQuantum1998a}. Practically, the unfolding process can be viewed as a
smoothing and rescaling procedure where the originally observed spectrum \(\lambda_1 \le \dots \le
\lambda_n\) is smoothly mapped to the unfolded spectrum \(e_1, \dots, e_n\), such that \(\sum d_i/n
\approx 1\), if \(d_i = e_{i+1} - e_i \). For most empirical data, the true smoothing function is
not known, and so must be approximated
(\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}). Typically, a polynomial is
chosen (\cite{abul-magdUnfoldingSpectrumChaotic2014}).

A number of statistics (or ``spectral observables''; \cite{mehtaRandomMatrices2004}) can then be
computed from this unfolded spectrum. In the literature, the two most common points of comparison
for these unfolded features are typically the features of the Poisson or Gaussian Diagonal Ensemble
(GDE) and the GOE (see e.g.
\cite{santhanamStatisticsAtmosphericCorrelations2001,jalanUncoveringRandomnessSuccess2014,bandyopadhyayUniversalityComplexNetworks2007,agrawalQuantifyingRandomnessProtein2014,raiRandomnessPreservedPatterns2015,
sebaRandomMatrixAnalysis2003,wangSpectralPropertiesTemporal2015,wangRandomMatrixTheory2016}). The
GOE (or GDE) consists of real, symmetric matrices where entries (or only diagonal entries) are
normal random variables with zero mean and unit variance. When the empirically observed features are
close to those of the GOE, this is usually interpreted as indicating a high degree of randomness and
statistical independence in the data, and when the observed features are closer to those expected in
the GDE or Poisson case, this is interpreted as indicating a high degree of coherence or dependence
(\cite{santhanamStatisticsAtmosphericCorrelations2001,jalanUncoveringRandomnessSuccess2014,bandyopadhyayUniversalityComplexNetworks2007,agrawalQuantifyingRandomnessProtein2014,raiRandomnessPreservedPatterns2015,
sebaRandomMatrixAnalysis2003,wangSpectralPropertiesTemporal2015,wangRandomMatrixTheory2016,matharooSpontaneousBackpainAlters2020}).

The Nearest-Neighbours Spacings Distribution (NNSD) is the most basic RMT feature, and is the
distribution of the spacings (differences) of adjacent unfolded eigenvalues. Normally, the NNSD is
analysed visually via a histogram of the distribution of spacings, however it is also possible to
analyse it more formally by fitting the Brody distribution (\cite{brodyStatisticalMeasureRepulsion1973}):

@@

Whereas the NNSD and Brody parameter characterize the local properties of eigenvalues close on the
spectrum, the long-range properties relating eigenvalues distant on the spectrum can be
characterized by both the spectral rigidity and level number variance
(\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}. The spectral rigidity,
\(\Delta_3(L)\), is calculated for any real value \(L\) as:

@@

where \(\eta(\lambda)\) is the number of unfolded eigenvalues less than or equal to \(\lambda\),
\(\langle \cdot \rangle_c\) denotes the average with respect to all starting points \(c\), and where
\(A\) and \(B\) denote the slope and intercept, respectively, of the least squares fit of a straight
line to \(\eta(\lambda)\) on \([c, c+L]\) \cite{guhrRandommatrixTheoriesQuantum1998a}. The level
number variance, \(\Sigma^2(L)\), is closely related to the spectral rigidity
(\cite{mehtaRandomMatrices2004}), and is calculated as:

@@

>>>>

\documentclass[NETN,manuscript]{stjour-new}
\articletype{Research}
\usepackage{graphicx}
\graphicspath{ {figs/} }
% \input{preamble.tex}
% \documentclass{article}
% \usepackage[utf8]{inputenc}
% \usepackage{geometry}
% \usepackage{amsmath}
% \usepackage{graphicx}
% \usepackage[style=apa]{biblatex}
% \usepackage{natbib}
% \usepackage[font=footnotesize,labelfont=bf]{caption}
% \addbibresource{rmt.bib}
% \bibliographystyle{plain}
% \bibliography{rmt}
\begin{document}

\title{Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations}
\author[Derek M. Berger, Gurpreet S. Matharoo, Jacob Levman]{
    Derek M. Berger\affil{1},
    Gurpreet S. Matharoo\affil{2}
    \and Jacob Levman\affil{3}
}
\affiliation{1}{Computer Science, St. Francis Xavier University, Antigonish, Canada}
\affiliation{2}{Physics, St. Francis Xavier University, Antigonish, Canada}
\affiliation{3}{Computer Science, St. Francis Xavier University, Antigonish, Canada}

\date{\today}

% \maketitle

% \keywords{random matrix theory, feature extraction, functional connectivity,
% functional magnetic resonance imaging, voxelwise connectivity, predictive analysis}

\begin{abstract}
    Previous studies have investigated the potential of using analytic techniques from Random Matrix
    Theory (RMT) to investigate magnetic resonance imaging (MRI) data. We assess the potential
    application of RMT-based features for the analysis of functional MRI (fMRI) across diverse

    datasets. As novel contributions, we (1) assess the potential for RMT-inspired, whole-brain
    features extracted from voxel-wise functional connectivity to contain information useful for
    classifying between various psychological processes, (2) assess these features’
    predictive—rather than explanatory—value, (3) investigate the effect of varying RMT analysis
    methods on the robustness of study findings, and (4) make code publicly available to extract
    these features from a wide variety of data. We find preliminary evidence suggesting that
    RMT-inspired features may have unique potential in analyses of fMRI functional connectivity.
\end{abstract}

@@

Whether in the presence of experimental stimuli, or the relative absence, as in a resting state,
complex functional connectivity networks are ubiquitous
\citep{bucknerBrainDefaultNetwork2008,foxCoverHumanBrain2005,gonzalez-castilloTaskbasedDynamicFunctional2018,hermundstadStructuralFoundationsRestingstate2013}.
This complexity suggests fMRI is a candidate to be studied using Random Matrix Theory (RMT), a set
of mathematical tools originally developed some 50 years ago to solve complex problems in nuclear
physics \citep{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}. In various
physical systems where nuclei can have different possible states, precisely modeling the entirety of
the interactions between nuclei is often intractable. For such large, complex systems, the
eigenvalues (or spectra) efficiently summarize the totality of the interactions between components
of the system, and RMT describes the expected behaviour of such eigenvalues. By analysing the
spectrum statistically, complex systems can be investigated by comparison to the universal
properties predicted by RMT.

\begin{sloppypar}
These universal properties have been observed in diverse phenomena. In small-scale systems, RMT
universalities have been observed in quantum chaotic systems, complex nuclei, atoms, molecules and
disordered mesoscopic systems \cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004,brodyRandommatrixPhysicsSpectrum1981,beenakkerRandommatrixTheoryQuantum1997,bohigasHigherOrderCorrelationsSpectra1985,wintgenLevelStatisticsQuantized1988,pandeySkewOrthogonalPolynomialsUniversality2001}, and at larger scales, RMT has been applied to atmospheric
physics \citep{santhanamStatisticsAtmosphericCorrelations2001}, stock cross-correlations \cite{plerouRandomMatrixApproach2002} , social networks \cite{jalanUncoveringRandomnessSuccess2014}, random networks \cite{bandyopadhyayUniversalityComplexNetworks2007},
network-formation in liquids \citep{sastrySpectralStatisticsInstantaneous2001,matharooSpectralStatisticsQuenched2009}  and amorphous clusters \citep{sarkarUniversalityVibrationalSpectra2004,matharooVibrationalSpectraAmorphous2005,matharooUniversalityVibrationalSpectra2008}. Within biological systems,
RMT has also been used to successfully model aspects of  amino acid functional relationships \citep{bhadolaTargetingFunctionalMotifs2016},
synchrony in epileptic seizures \citep{osorioPhasesynchronizationRandommatrixBased2011}, and in protein-protein interactions both in different species \citep{agrawalQuantifyingRandomnessProtein2014} and breast cancer \citep{raiRandomnessPreservedPatterns2015}.
\end{sloppypar}

In recent years, RMT analyses have been applied to study brain functioning.  The earliest  study
demonstrated that spectra of the correlations between electroencephalographic (EEG) signals closely
resemble those of the Gaussian Orthogonal Ensemble (GOE) \citep{sebaRandomMatrixAnalysis2003}.
Somewhat more recently, RMT was used to evaluate the quality of whole brain features extracted from
fMRI data \citep{voultsidouFeatureEvaluationFMRI2007,verganiRestingStateFMRI2019}. RMT has also
been used in diffusion MRI to aid in the selection of the number of components to employ in
principal-component reduction analysis and denoising
\citep{veraartDenoisingDiffusionMRI2016,verganiRestingStateFMRI2019,ulfarssonDimensionEstimationNoisy2008}.

Finally, RMT has been used in ROI-based fMRI functional-connectivity studies to investigate
differences between rest and task states \citep{wangSpectralPropertiesTemporal2015}, between
subjects with and without ADHD \citep{wangRandomMatrixTheory2016}, and between pain and non-pain
states \citep{matharooSpontaneousBackpainAlters2020}. Across these three studies, the spectra of
resting or low-attention states exhibited properties closest to the GOE.

@@

However, since previous studies employing RMT for feature extraction
\citep{wangRandomMatrixTheory2016,matharooSpontaneousBackpainAlters2020} have partially interpreted
their findings with respect to attentional processes, we have elected to use datasets that involve
attention in some way. This includes measures of attention or vigilance, or comparisons between a
condition involving more attention and focus (e.g. a task state) or conditions which should involve
considerably less attention or focus (e.g. resting-state).

@@

Dataset scan parameters and acquisition details are summarized in Table \ref{table:1}, and sample
and subgroup sizes are summarized in Table \ref{table:2}. For all datasets, we use the raw
(unpreprocessed) scans, and perform our own preprocessing (detailed later). Given the complexity of
these datasets, we highlight below only those aspects relevant to the current paper.

@@

\subsubsection{Dataset 1}
Dataset 1 (referred to in figures and tables with the label ``PSYCH'') has detailed and extensive
information available in \cite{gorgolewskiHighResolution7Tesla2015}. Subjects completed a battery of
psychological measurements, and various resting-state fMRI (rs-fMRI) images were obtained. Only the
whole-brain 7-Tesla rs-fMRI scans were used for the current paper, to make the results comparable to
the other included datasets. For the purpose of keeping with the ``attention'' theme of the analyses
in this paper, three pairings of subgroups were constructed via z-score splits of the various
sum-scores of scale items involving concepts closely related to attention.

\paragraph{Trait Attention}
The PANAS-X \citep{watsonPANASXManualPositive1994} is a 9-point Likert-type self-report measure
that queries the extent to which subjects agree particular affective terms have applied to them over
a recent time period (e.g. a few weeks). An ad hoc (i.e. face-valid) sum-score was constructed from
the scale items: ``attentive'', ``alert'', ``interested'', ``concentrating'', and reverse-scored items
``sluggish'', ``tired'', ``sleepy'', ``drowsy''. Subjects with a z-score greater / less than zero on this
sum-score were placed in the high / low ``trait attention'' category

\paragraph{Task Attention}
The Conjunctive Continuous Performance Task (CCPT;
\cite{shalevConjunctiveContinuousPerformance2011} is a behavioural measure of sustained attention
where subjects must react only if they see a red square among other different coloured figures. For
this paper, a sum score was constructed by subtracting the number of false positive and false
negative responses from the number of success responses, and subjects with a z-score greater / less
than zero on this scale were placed in high / low ``task attention'' category.

\paragraph{Vigilance}
The Mini New York cognition questionnaire assesses various elements of the subject's cognitive
state, e.g. ``my thoughts involved other people'', ``I thought about something positive'' (translated
from German; \cite{gorgolewskiCorrespondenceIndividualDifferences2014}. One question assesses
alertness (``I was fully awake'' / ``war ich vollkommen wach'') and was used to create high / low
``vigilance'' conditions in the same manner as the previous two splits.

\subsubsection{Dataset 2}
Dataset 2 (referred to in Fig. and tables with the label ``OSTEO'') is detailed in
\cite{tetreaultBrainConnectivityPredicts2016}. Subjects included healthy controls (``nopain''
condition) and individuals with knee osteoarthritis (``pain'' condition) treated for 2 weeks with
either placebo (``pain'' condition) or duloxetine (``duloxetine'' condition). An additional grouping was
also created with both the placebo and duloxetine data (``allpain'' condition), allowing for six
subgroup comparisons. All subjects were imaged with whole-brain rs-fMRI. Scan parameters are
documented in Table \ref{table:1}. The use of rs-fMRI and RMT for examining pain makes this dataset
complementary to \cite{matharooSpontaneousBackpainAlters2020}.

\subsubsection{Dataset 3}
Dataset 3  (referred to here with the label ``REFLECT'', due to the ``reflective” nature of the
cognitive tasks) is described in full detail by \cite{dupreMultiechoFMRIReplication2016}. Both
multi-echo rs- and task-fMRI scans were conducted with scan parameters summarized in Table \ref{table:2}. Tasks
involved viewing various affective images \citep[see][for details]{sprengPatternsBrainActivity2010}
and performing tasks related to autobiographical memory, prospection, or theory-of-mind reasoning.
Continuing with the theme of broadly comparing high versus low attention throughout this paper, the
different tasks were simply treated generically as ``attending'', in contrast to the non-attentive
resting-state scans. This resulted in four task state scans and one resting-state scan per subject.

In order to combine the multi-echo data into a single, usable matrix, two methods were employed. In
the ``summed'' condition, echoes were combined via summation, a method which has value in providing an
image with overall increased contrast \citep{kunduMultiechoFMRIReview2017}. In the ``interleave''
condition, echoes were interleaved at each voxel: that is, if we denote the voxel's ith signal value
from echo j as , then the interleaved voxel of the new image has the time-series \([t_{1,1},
t_{2,1}, t_{3,1}, \dots, t_{1,i}, t_{2,i}, t_{3,i}, \dots, t_{1,n}, t_{2,n}, t_{3,n}]\).

@@

\subsubsection{Dataset 4}
Dataset 4 (referred to in Fig. and tables with the label ``PARK'' or ``PARKINSONS'') is documented and
described in detail in \cite{madhyasthaDynamicConnectivityRest2015}. Scan parameters are documented
in Table \ref{table:1}. Subjects with non-demented Parkinson’s Disease (PD) and healthy controls performed a
number of repetitions of the Attention Network Task
\citep[ANT;][]{fanActivationAttentionalNetworks2005}. Subgroup details are available in Table \ref{table:1}. As the ANT
task demands high attention, it is not as straightforward to create ``high'' or ``low'' attention
groups. Instead, since attention presumably is high and varies little in this dataset, and subgroups
instead differ by virtue of the presence or absence of PD, any results here serve as a useful
comparison point for the results from studies that should clearly involve attentional differences.
This dataset also included extensively preprocessed scans, which were included in analyses to
examine the effects of preprocessing on the extracted RMT features.

\subsubsection{Dataset 5}
Dataset 5 (referred to in Fig. and tables with the label ``LEARN'') is documented and described in
detail in \cite{schapiroHumanHippocampalReplay2018}. In this study, both rs-fMRI and task fMRI scans
were available for all subjects, allowing for within-subject comparisons. The complex task involved
multiple phases, wherein subjects were exposed to named, black-and-white outlined satellite images
related by family-resemblance of the components \citep[see figure 1
of][]{schapiroHumanHippocampalReplay2018}. After this learning task, subjects were tested on their
memory of the various learned features. Scans corresponding to both the initial learning and
subsequent testing phase were included in the high-attention ``task'' subgroup for the purposes of
the analysis of this paper. Resting state scans were included in the low-attention ``rest''
subgroup.

\subsection{Preprocessing}
Some consideration should be given to whether it is appropriate to apply standard preprocessing
steps before extracting RMT features, as preprocessing steps should be developed, tested, and
employed always with respect to a set of inferential goals and a specific analytic framework.
However, these preprocessing steps are increasingly recognized to have complex effects on fMRI
analyses \citep{parkerBenefitSliceTiming2019}. For example, spatial smoothing has clear benefits
when employing a general linear model
\citep{miklEffectsSpatialSmoothing2008,alakorkkoEffectsSpatialSmoothing2017,worsleyAnalysisFMRITimeSeries1995},
but is less motivated in functional network analyses with small ROIs, where the effect of smoothing
can be considerable
\citep{alakorkkoEffectsSpatialSmoothing2017,goelmanMaximizingNegativeCorrelations2014,trianaEffectsSpatialSmoothing2020},
with these effects interacting complexly with factors like field strength and image resolution
\citep{triantafyllouEffectSpatialSmoothing2006}. Similarly, image registration is usually required
to allow comparisons across (or even within) subjects, but the choice of registration strategy can
have dramatic effects on functional network analyses \citep{magalhaesImpactNormalizationSegmentation2014}.

@@

For this paper’s analyses, brain extraction is necessary to limit analyses to the appropriate
voxels, motion correction is necessary to allow proper brain extraction, and both motion-correction
and slice-time correction simplify the interpretation of voxel-voxel correlations across voxel
locations. Whereas physiological noise regression and bias field correction would perhaps be
appropriate here, there is insufficient data to perform these steps for all datasets (Table \ref{table:2}).
Thus, motion-correction was performed with FSL's MCFLIRT
\citep{jenkinsonImprovedOptimizationRobust2002}, slicetime correction with the default method in
FSL, and brain extraction was performed with BET \citep{smithFastRobustAutomated2002}. Due to
differences in the size of the REFLECT data, the higher-resolution resting-state data was
interpolated to have the same spatial dimensions as the task data.

@@

Two classes of features were examined in this study: those that require unfolding, and those that do
not. In order to compare the observed eigenvalue distribution to some theoretical predictions, the
eigenvalues must be unfolded \citep{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}.
The details and motivation behind the unfolding procedure are well documented in
\cite{guhrRandommatrixTheoriesQuantum1998a}. Practically, the unfolding process can be viewed as a
smoothing and rescaling procedure where the originally observed spectrum \(\lambda_1 \le \dots \le
\lambda_n\) is smoothly mapped to the unfolded spectrum \(e_1, \dots, e_n\), such that \(\sum d_i/n
\approx 1\), if \(d_i = e_{i+1} - e_i \). For most empirical data, the true smoothing function is
not known, and so must be approximated
\citep{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}. Typically, a polynomial is
chosen \citep{abul-magdUnfoldingSpectrumChaotic2014}.

A number of statistics or ``spectral observables'' \citep{mehtaRandomMatrices2004} can then be
computed from this unfolded spectrum. In the literature, the two most common points of comparison
for these unfolded features are typically the features of the Poisson or Gaussian Diagonal Ensemble
(GDE) and the GOE
\cite[see e.g.][]{santhanamStatisticsAtmosphericCorrelations2001,jalanUncoveringRandomnessSuccess2014,bandyopadhyayUniversalityComplexNetworks2007,agrawalQuantifyingRandomnessProtein2014,raiRandomnessPreservedPatterns2015,
sebaRandomMatrixAnalysis2003,wangSpectralPropertiesTemporal2015,wangRandomMatrixTheory2016}. The
GOE (or GDE) consists of real, symmetric matrices where entries (or only diagonal entries) are
normal random variables with zero mean and unit variance. When the empirically observed features are
close to those of the GOE, this is usually interpreted as indicating a high degree of randomness and
statistical independence in the data, and when the observed features are closer to those expected in
the GDE or Poisson case, this is interpreted as indicating a high degree of coherence or dependence
\citep{santhanamStatisticsAtmosphericCorrelations2001,jalanUncoveringRandomnessSuccess2014,bandyopadhyayUniversalityComplexNetworks2007,agrawalQuantifyingRandomnessProtein2014,raiRandomnessPreservedPatterns2015,
sebaRandomMatrixAnalysis2003,wangSpectralPropertiesTemporal2015,wangRandomMatrixTheory2016,matharooSpontaneousBackpainAlters2020}.

The Nearest-Neighbours Spacings Distribution (NNSD) is the most basic RMT feature, and is the
distribution of the spacings (differences) of adjacent unfolded eigenvalues. Normally, the NNSD is
analysed visually via a histogram of the distribution of spacings, however it is also possible to
analyse it more formally by fitting the Brody distribution \citep{brodyStatisticalMeasureRepulsion1973}:

@@

Whereas the NNSD and Brody parameter characterize the local properties of eigenvalues close on the
spectrum, the long-range properties relating eigenvalues distant on the spectrum can be
characterized by both the spectral rigidity and level number variance
\citep{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}. The spectral rigidity,
\(\Delta_3(L)\), is calculated for any real value \(L\) as:

@@

where \(\eta(\lambda)\) is the number of unfolded eigenvalues less than or equal to \(\lambda\),
\(\langle \cdot \rangle_c\) denotes the average with respect to all starting points \(c\), and where
\(A\) and \(B\) denote the slope and intercept, respectively, of the least squares fit of a straight
line to \(\eta(\lambda)\) on \([c, c+L]\) \citep{guhrRandommatrixTheoriesQuantum1998a}. The level
number variance, \(\Sigma^2(L)\), is closely related to the spectral rigidity
\citep{mehtaRandomMatrices2004}, and is calculated as:

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=\textwidth]{figure1}} \caption{
    (Colour online) Spectral rigidity for controls (nopain) vs. those with osteopathic pain taking
    duloxetine, trimming largest eigenvalue only. Solid lines indicate group mean rigidity, and
    shaded regions correspond to 99\% percentile bootstrapped intervals for each group. Vertical
    axes vary to better depict group-wise overlap.}
\label{fig:rigidity}
\end{figure}
